---
title: "p8105_hw2_sg3891"
author: Sandya Ganesh
date: 2021-10-06
output: github_document
---

```{r setup, echo = FALSE, message = FALSE}
library(tidyverse)
library(readxl)
```
## Question 1

#### Read and clean the Mr. Trash Wheel sheet

We will read the relevant excel sheet and omit non-data entries and rows that do not include dumpster-specific data. Additionally, we will round the number of sports balls to the nearest integer.

```{r trash}
trashwheel_df = 
  read_excel("data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx", sheet = 1, range = "A2:N534") %>% 
  janitor::clean_names() %>% 
  drop_na(dumpster) %>% 
  mutate(
    sports_balls = round(sports_balls, 0))
```

#### Read and clean the precipitation data for 2019 and 2018

In the code chunk below, we read and clean precipitation data for 2018 and 2019 in 2 separate data frames. We will omit rows without precipitation data, add a variable for year, and finally combine the precipitation data sets.

```{r precipitation}
p2019_df = 
  read_excel("data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx", 
             sheet = 6, 
             range = "A2:B14") %>% 
  janitor::clean_names() %>% 
  mutate(
    year = 2019)

p2018_df = 
  read_excel("data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx", sheet = 7, range = "A2:B14") %>% 
  janitor::clean_names() %>% 
  mutate(
    year = 2018)

pcombined_df =
  bind_rows(p2019_df, p2018_df) %>%
  mutate(month = month.name[month])
```

#### Description of Data
The number of observations in the Mr. Trash Wheel sheet is `r nrow(trashwheel_df)` observations with `r ncol(trashwheel_df)` variables. Some key variables in this data set include `r names(trashwheel_df %>% select(c(1, 3, 5)))`. The dumpsters are numbered from `r  min(pull(trashwheel_df, dumpster))` to `r  max(pull(trashwheel_df, dumpster))`. Additionally, this data is collected from `r  min(pull(trashwheel_df, year))` to `r  max (pull(trashwheel_df, year))`. The total weight of the trash in the data set is `r  sum(pull(trashwheel_df, weight_tons))` tons. The median number of sports balls in a dumpster in 2019 is `r median(pull(filter(trashwheel_df, year == 2019), sports_balls))` balls.

The total number of observations in the precipitation data set is `r nrow(pcombined_df)` observations with `r ncol(pcombined_df)` variables. Some key variables in this data set include `r names(trashwheel_df %>% select(c(1, 2, 3)))`. Average total rainfall in 2018 is `r mean(pull(filter(pcombined_df, year == 2018), total))` inches and `r sum(pull(filter(pcombined_df, year == 2019), total))` inches in 2019. The total precipitation is 2018 was `r sum(pull(filter(pcombined_df, year == 2018), total))` inches and the total precipitation in 2019 was `r sum(pull(filter(pcombined_df, year == 2019), total))` inches.

## Question 2

Our goal for this question is to merge 3 csv files into a single data frame.

#### Read and clean the pols-month.csv file

In addition to cleaning the dataset, the code below creates a new variable called president which provides information on the political party of the president at the given time and month (gop for republican, dem for democratic).

```{r pols, message = FALSE}
pols_month_df = 
  read_csv("data/pols-month.csv") %>% 
  janitor::clean_names() %>%
  separate(mon,into = c("year","month","day"),sep = "-") %>%
  pivot_longer(
    c(prez_gop,prez_dem), names_prefix = "prez_", names_to = "president") %>% 
  filter(value > 0) %>%
  select(-day, -value) %>% 
  mutate(
    year = as.integer(year),
    month = as.integer(month), 
    month = month.name[month]) %>% 
  arrange(year, month)

```

#### Read and clean the snp.csv file

```{r snp, message = FALSE}
snp_df = 
  read_csv("data/snp.csv") %>% 
  janitor::clean_names() %>%
  separate(date,into = c("month","day","year"),sep = "/") %>%
  mutate(
    year = as.integer(year),
    year = ifelse(year > 21, 1900 + year, 2000 + year),
    month = as.integer(month), 
    month = month.name[month]) %>% 
  select(-day) %>% 
  select(year, month, close) %>% 
  arrange(year, month)

```

#### Read and clean the unemployment.csv file

```{r unemp, message = FALSE}
unemp_df = 
  read_csv("data/unemployment.csv") %>% 
  janitor::clean_names() %>%
  pivot_longer(
    c(jan:dec),
    names_to = "month", 
    values_to = "unemployment") %>%
  mutate(
    year = as.integer(year),
    month = recode(month, "jan" = "January", "feb" = "February", "mar" = "March", "apr" = "April",
                   "may" = "May", "jun" = "June", "jul" = "July", "aug" = "August", "sep" = 
                    "September", "oct" = "October", "nov" = "November", "dec" = "December")) %>% 
  arrange(year, month)

```

#### Merge and describe data
```{r merge, message = FALSE}
p2_merge_df =  left_join(pols_month_df, snp_df, by = c("year", "month"))

p2_final_df = left_join(p2_merge_df, unemp_df, by = c("year", "month"))

```

The pols_month_df dataset contains `r nrow(pols_month_df)` observations of `r ncol(pols_month_df)` variables, related to the number of national politicians who are democratic or republican at any given time. It includes the variables `r names(pols_month_df)`. The newly created president variable in our dataset indicates whether the president in office in the given month and year was republican (gop) or democratic (dem).

The snp_df dataset contains `r nrow(snp_df)` observations of `r ncol(snp_df)` variables related to Standard & Poor’s stock market index (S&P), often used as a representative measure of stock market as a whole. It includes the variables `r names(snp_df)`.

The unemp_df dataset contains `r nrow(unemp_df)` observations of `r ncol(unemp_df)` variables. It includes the variables `r names(unemp_df)`, and provides the percentage of unemployment in each month of the associated year.

The merged data set, called p2_final_df, contains `r nrow(p2_final_df)` observations of `r ncol(p2_final_df)` variables. The range of years is from `r  min(pull(p2_final_df, year))` to `r  max(pull(p2_final_df, year))`. The names of key variables include `r names(p2_final_df %>% select(c(1, 9, 10, 11)))`. From these key variables, we are able to identify the year and presidential party for each observation. Additionally, the "close" variable indicates the closing values of the S&P stock index on the associated month and year, and the unemployment variable indicates the percentage of unemployment in the given month and year.

## Question 3

#### Read and clean the NYC open data on Popular_Baby_names.csv

The code below cleans the data in the ethnicity column, as there are records coded as "ASIAN AND PACI" and "ASIAN AND PACIFIC ISLANDER." Additionally HISPANIC is coded as HISP in some rows, so we need to create a consistent coding within the data set.

```{r nyc_open, message = FALSE}
child_names_df = 
  read_csv("data/Popular_Baby_Names.csv") %>% 
  janitor::clean_names() %>% 
  mutate(
    childs_first_name = str_to_title(childs_first_name),
    ethnicity = recode(ethnicity,
    "ASIAN AND PACI" = "ASIAN AND PACIFIC ISLANDER",
    "BLACK NON HISP" = "BLACK NON HISPANIC", 
    "WHITE NON HISP" = "WHITE NON HISPANIC")) %>%
  distinct() # use this to ensure that each row is unique
```

#### Tables for popular baby names over time

The code chunk below produces a well-structured, reader-friendly table showing the rank in popularity of the name “Olivia” as a female baby name over time.


```{r olivia, message = FALSE}
olivia_df = 
  filter(
    child_names_df, childs_first_name == "Olivia" & gender == "FEMALE") %>% 
  select(year_of_birth, ethnicity, rank) %>%
    pivot_wider(
      names_from = "year_of_birth",
      values_from = "rank")

knitr::kable(olivia_df)
```

The code chunk below produces a well-structured, reader-friendly table showing the rank in popularity of the most popular name as a male baby name over time.

```{r male, message = FALSE}
male_df = 
  filter(
    child_names_df, gender == "MALE" & rank == 1) %>% 
  select(year_of_birth, ethnicity, childs_first_name) %>%
    pivot_wider(
      names_from = "year_of_birth",
      values_from = "childs_first_name")

knitr::kable(male_df)
```

#### Scatter plot 

The code chunk below produces a scatter plot showing the number of children with a name (y axis) against the rank in popularity of that name (x axis), for male, white, non hispanic children born
in 2016.

```{r scatter, message = FALSE}
scatter_df = 
  filter(
    child_names_df,
    gender == "MALE",
    ethnicity == "WHITE NON HISPANIC",
    year_of_birth == 2016)


ggplot(scatter_df, aes(x = rank, y = count)) + geom_point()
```

