---
title: "p8105_hw2_sg3891"
author: Sandya Ganesh
date: 2021-10-06
output: github_document
---

```{r setup, echo = FALSE, message = FALSE}
library(tidyverse)
library(readxl)

```
## Question 1

#### Read and clean the Mr. Trash Wheel sheet

```{r trash}
trashwheel_df = 
  read_excel("data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx", sheet = 1, range = "A2:N534") %>% 
  janitor::clean_names() %>% 
  drop_na(dumpster) %>% 
  mutate(
    sports_balls = round(sports_balls, 0))

```

#### Read and clean the precipitation data for 2019 and 2018
```{r precipitation}
p2019_df = 
  read_excel("data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx", 
             sheet = 6, 
             range = "A2:B14") %>% 
  janitor::clean_names() %>% 
  mutate(
    year = 2019)

p2018_df = 
  read_excel("data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx", sheet = 7, range = "A2:B14") %>% 
  janitor::clean_names() %>% 
  mutate(
    year = 2018)

pcombined_df =
  bind_rows(p2019_df, p2018_df) %>%
  mutate(month = month.name[month])

```

#### Description of Data
The number of observations in the Mr. Trash Wheel sheet is `r nrow(trashwheel_df)` observations with `r ncol(trashwheel_df)` variables. Some key variables in this data set include `r names(trashwheel_df %>% select(c(1, 3, 5)))`. The dumpsters are numbered from `r  min(pull(trashwheel_df, dumpster))` to `r  max(pull(trashwheel_df, dumpster))`. Additionally, this data is collected from `r  min(pull(trashwheel_df, year))` to `r  max (pull(trashwheel_df, year))`. The total weight of the trash in the data set is `r  sum(pull(trashwheel_df, weight_tons))` tons. The median number of sports balls in a dumpster in 2019 is `r median(pull(filter(trashwheel_df, year == 2019), sports_balls))` balls.

The total number of observations in the precipitation data set is `r nrow(pcombined_df)` observations with `r ncol(pcombined_df)` variables. Some key variables in this data set include `r names(trashwheel_df %>% select(c(1, 2, 3)))`. Average total rainfall in 2018 is `r mean(pull(filter(pcombined_df, year == 2018), total))` inches and `r sum(pull(filter(pcombined_df, year == 2019), total))` inches in 2019. The total precipitation is 2018 was `r sum(pull(filter(pcombined_df, year == 2018), total))` inches and the total precipitation in 2019 was `r sum(pull(filter(pcombined_df, year == 2019), total))` inches.

## Question 2

#### Read and clean the pols-month.csv file

```{r pols, message = FALSE}
pols_month_df = 
  read_csv("data/pols-month.csv") %>% 
  janitor::clean_names() %>%
  separate(mon,into = c("year","month","day"),sep = "-") %>%
  pivot_longer(
    c(prez_gop,prez_dem), names_prefix = "prez_", names_to = "president") %>% 
  filter(value > 0) %>%
  select(-day, -value) %>% 
  mutate(
    year = as.integer(year),
    month = as.integer(month), 
    month = month.name[month]) %>% 
  arrange(year, month)

```

#### Read and clean the snp.csv file

```{r snp, message = FALSE}
snp_df = 
  read_csv("data/snp.csv") %>% 
  janitor::clean_names() %>%
  separate(date,into = c("month","day","year"),sep = "/") %>%
  mutate(
    year = as.integer(year),
    year = ifelse(year > 21, 1900 + year, 2000 + year),
    month = as.integer(month), 
    month = month.name[month]) %>% 
  select(-day) %>% 
  select(year, month, close) %>% 
  arrange(year, month)

```

#### Read and clean the unemployment.csv file

```{r unemp, message = FALSE}
unemp_df = 
  read_csv("data/unemployment.csv") %>% 
  janitor::clean_names() %>%
  pivot_longer(
    c(jan:dec),
    names_to = "month", 
    values_to = "unemployment") %>% 
  mutate(
    year = as.integer(year),
    month = recode(month, "jan" = "January", "feb" = "February", "mar" = "March", "apr" = "April", "may" = "May", "jun" = "June", "jul" = "July", "aug" = "August", "sept" = "September", "oct" = "October", "nov" = "November", "dec" = "December")) %>% 
  arrange(year, month)

```

#### Merge and describe data
```{r merge, message = FALSE}
p2_merge_df =  left_join(pols_month_df, snp_df, by = c("year", "month"))

p2_final_df = left_join(p2_merge_df, unemp_df, by = c("year", "month"))

```

The pols_month_df dataset contains `r nrow(pols_month_df)` observations of `r ncol(pols_month_df)` variables, related to the number of national politicians who are democratic or republican at any given time.

The snp_df dataset contains `r nrow(snp_df)` observations of `r ncol(snp_df)` variables related to Standard & Poor’s stock market index (S&P), often used as a representative measure of stock market as a whole. 

The merged dataset, called p2_final_df, contains `r nrow(p2_final_df)` observations of `r ncol(p2_final_df)` variables. The range of years is from `r  min(pull(p2_final_df, year))` to `r  max(pull(p2_final_df, year))`. The names of key variables include `r names(p2_final_df %>% select(c(1, 9, 10, 11)))`. From these key variables, we are able to identify the year and president for each observation. Additionally, the close variable indicates the closing values of the S&P stock index on the associated month and year, and the unemployment variable indicates the percentage of unemployment in the given month and year.

## Question 3

#### Read and clean the NYC open data on Popular_Baby_names.csv

We need to clean up the data in the ethnicity column, as I see records coded as "ASIAN AND PACI" and "ASIAN AND PACIFIC ISLANDER." Additionally HISPANIC is coded as HISP in some rows, so we need to create a consistent coding within the dataset.


```{r nyc_open, message = FALSE}
child_names_df = 
  read_csv("data/Popular_Baby_Names.csv") %>% 
  janitor::clean_names() %>% 
  mutate(
    childs_first_name = str_to_title(childs_first_name),
  "ASIAN AND PACI" = "ASIAN AND PACIFIC ISLANDER",
  "BLACK NON HISP" = "BLACK NON HISPANIC", 
  "WHITE NON HISP" = "WHITE NON HISPANIC") %>%
  distinct()
```


#### Table for popular baby name over time

Produce a well-structured, reader-friendly table showing the rank in popularity of the name “Olivia” as a female baby name over time.


```{r olivia, message = FALSE}
olivia_df = 
  filter(
    child_names_df, childs_first_name == "Olivia" & gender == "FEMALE") %>% 
  select(year_of_birth, ethnicity, rank) %>%
    pivot_wider(
      names_from = "year_of_birth",
      values_from = "rank")

knitr::kable(olivia_df)
```

Produce a well-structured, reader-friendly table showing the rank in popularity of the most popular name as a male baby name over time.


```{r mmale, message = FALSE}
male_df = 
  filter(
    child_names_df, gender == "MALE" & rank == 1) %>% 
  select(year_of_birth, ethnicity, childs_first_name) %>%
    pivot_wider(
      names_from = "year_of_birth",
      values_from = "childs_first_name")

knitr::kable(male_df)
```

