---
title: "p8105_hw2_sg3891"
author: Sandya Ganesh
date: 2021-10-06
output: github_document
---

```{r setup, echo = FALSE, message = FALSE}
library(tidyverse)
library(readxl)
```
## Question 1

#### Read and clean the Mr. Trash Wheel sheet

```{r trash}
trashwheel_df = 
  read_excel("data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx", sheet = 1, range = "A2:N534") %>% 
  janitor::clean_names() %>% 
  drop_na(dumpster) %>% 
  mutate(
    sports_balls = round(sports_balls, 0))
```

#### Read and clean the precipitation data for 2019 and 2018
```{r precipitation}
p2019_df = 
  read_excel("data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx", sheet = 6, range = "A2:B14") %>% 
  janitor::clean_names() %>% 
  mutate(
    year = 2019)

p2018_df = 
  read_excel("data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx", sheet = 7, range = "A2:B14") %>% 
  janitor::clean_names() %>% 
  mutate(
    year = 2018)

pcombined_df =
  bind_rows(p2019_df, p2018_df) %>%
  mutate(month = month.name[month])
```

#### Description of Data
The number of observations in the Mr. Trash Wheel sheet is `r nrow(trashwheel_df)` observations and the data set also contains `r ncol(trashwheel_df)` columns/variables. Some key variables in this data set include `r names(trashwheel_df %>% select(c(1, 3, 5)))`. The dumpsters are numbered from `r  min(pull(trashwheel_df, dumpster))` to `r  max(pull(trashwheel_df, dumpster))`. Additionally, this data is collected from `r  min(pull(trashwheel_df, year))` to `r  max (pull(trashwheel_df, year))`. The total weight of the trash in the data set is `r  sum(pull(trashwheel_df, weight_tons))` tons. The median number of sports balls in a dumpster in 2019 is `r median(pull(filter(trashwheel_df, year == 2019), sports_balls))` balls.

The total number of observations in the precipitation data set is `r nrow(pcombined_df)` observations and the dataset also includes `r ncol(pcombined_df)` columns/variables. Some key variables in this data set include `r names(trashwheel_df %>% select(c(1, 2, 3)))`. Average total rainfall in 2018 is `r mean(pull(filter(pcombined_df, year == 2018), total))` inches and `r sum(pull(filter(pcombined_df, year == 2019), total))` inches in 2019. The total precipitation is 2018 was `r sum(pull(filter(pcombined_df, year == 2018), total))` inches and the total precipitation in 2019 was `r sum(pull(filter(pcombined_df, year == 2019), total))` inches.

